---
title: "Getting Started with volvelle"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with volvelle}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(volvelle)
```

## Overview

**volvelle** provides a config-driven ROLLUP aggregation framework with an embeddable HTML drill-down widget. The name references the medieval rotating paper calculation instrument — nested analytical layers, each a distinct _face_ or facet of the data.

The core workflow is:

1. Supply a data frame and a YAML config file
2. `volvelle()` returns a structured list of ROLLUP-aggregated data frames — one per filter facet plus one unfiltered (`"full"`)
3. `volvelle_widget()` wraps the result in a self-contained interactive HTML widget

## The Demo Dataset

The package ships with `credit_portfolio`, a synthetic dataset of 500 credit exposures:

```{r}
head(credit_portfolio)
str(credit_portfolio)
```

## The Config File

The demo config at `inst/extdata/demo_config.yaml` declares:

- **Hierarchy**: `segment → sub_segment → counterparty`
- **Measures**: total exposure (sum), weighted-average PD and LGD, distinct obligor count
- **Derived field**: expected loss (post-aggregation product)
- **Filters**: performing, non-performing, large exposure

```{r}
cfg_path <- system.file("extdata", "demo_config.yaml", package = "volvelle")
cfg      <- parse_config(cfg_path)
cat(yaml::as.yaml(cfg))
```

## Running volvelle()

```{r}
result <- volvelle(credit_portfolio, cfg_path)
print(result)
```

The result is a named list of `data.table` objects, one per facet. The `"full"` facet always contains the unfiltered ROLLUP:

```{r}
# Show first 20 rows of the full facet
head(result$full, 20)
```

Metadata columns:

- `.id`: grouping-set index (1 = finest, N+1 = grand total)
- `.rollup_depth`: 0 = grand total, `length(hierarchy)` = leaf level

### Grand Total Row

```{r}
grand <- result$full[result$full$.rollup_depth == 0L, ]
grand[, c("exposure", "n_obligors", "expected_loss")]
```

## Interactive Widget

### Light Theme

```{r, out.width = "100%"}
volvelle_widget(result, title = "Credit Portfolio ROLLUP", theme = "light")
```

### Dark Theme

```{r, out.width = "100%"}
volvelle_widget(result, title = "Credit Portfolio ROLLUP", theme = "dark")
```

## Accessing a Single Facet

You can extract any facet as a plain `data.frame` for downstream use:

```{r}
# Extract the 'performing' facet
performing_df <- as.data.frame(result, facet = "performing")
class(performing_df)
nrow(performing_df)

# Leaf-level rows only (finest granularity)
leaf <- performing_df[performing_df$.rollup_depth == max(performing_df$.rollup_depth), ]
head(leaf[, c("segment", "sub_segment", "counterparty", "exposure", "avg_pd")])
```

## Custom Filters (Programmatic Injection)

You can inject filters at call time via `extra_filters`, without modifying the config file. The filter structure mirrors the YAML `filters` block:

```{r}
extra <- list(
  watch_only = list(col = "status", op = "==", val = "watch")
)

result2 <- volvelle(credit_portfolio, cfg_path, extra_filters = extra)
facets(result2)
```

Alternatively, pre-filter the data manually before calling `volvelle()`:

```{r}
sovereign_data <- credit_portfolio[credit_portfolio$segment == "Sovereign", ]
result_sov     <- volvelle(sovereign_data, cfg_path)
print(result_sov)
```

## Saving the Widget to HTML

```{r, eval = FALSE}
w <- volvelle_widget(result, title = "Credit Portfolio")
htmlwidgets::saveWidget(w, "portfolio_report.html", selfcontained = TRUE)
```

The resulting `.html` file is fully self-contained and requires no external dependencies to view.
